---
title: "Mock-Tests in R using `mockery`"
author: "Russ Hyde"
date: "25 September 2018"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

```{r}
set.seed(1)

pkgs <- c(
  "ComplexHeatmap",
  "magrittr",
  "tidyr",
  "tibble",
  "testthat",
  "mockery",
  "functionalheatmap"
)

for (pkg in pkgs) {
  suppressPackageStartupMessages(
    library(pkg, character.only = TRUE)
  )
}
```

# Introduction

This document is mostly a recollection aid written while working on a package
called [`functionalheatmap`](https://github.com/russHyde/functionalheatmap).

While writing that package, some tests were required that were somewhat out
of my comfort zone. They involved redefining an R function at test-time using
the `mockery` package. I'll explain why it was necessary to do this after
a brief intro to my unfinished (and, some might say, superfluous) package.

# Basic use of `functionalheatmap`

I've been writing a wrapper around the
[
  `ComplexHeatmap`
](
  https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html
) package that allows the production of heatmaps using incremental steps and
requires less housekeeping than does the, otherwise superb, `ComplexHeatmap`
package.

Here's some example data for use in a heatmap:

```{r}
body_data <- matrix(
  rnorm(20),
  nrow = 5, dimnames = list(letters[1:5], LETTERS[1:4])
)

row_data <- data.frame(
  foo = c(FALSE, TRUE, FALSE, FALSE, TRUE),
  bar = 1:5,
  row.names = letters[1:5]
)
```

```{r}
as.data.frame(body_data)
```

```{r}
row_data
```

Given the above dataset, to produce a heatmap in ComplexHeatmap that had a
set of row-annotations (my usual use-case) you'd do something like the
following.

```{r}
my_heatmap <- ComplexHeatmap::Heatmap(
  body_data,
  cluster_columns = FALSE
)

my_annots <- ComplexHeatmap::HeatmapAnnotation(
  row_data,
  which = "row", show_annotation_name = TRUE
)
```

```{r, fig.width = 4, fig.height = 4.8}
my_heatmap + my_annots
```

A given row of your annotations has to match the corresponding row of the
heatmap's body.

The package `functionalheatmap` provides some wrapper functions to ensure that
the rows of the body and annotations match-up. This is helpful if you want to
make multiple related heatmaps based on different subsets of the main body-data
matrix or subsets of the annotations.

To plot a sub-heatmap for just those features where `foo` is false in
ComplexHeatmap you would do the following:

```{r}
# housekeeping variable
keep_rows <- which(!row_data$foo)

new_heatmap <- ComplexHeatmap::Heatmap(
  body_data[keep_rows, ],
  cluster_columns = FALSE
)

new_annots <- ComplexHeatmap::HeatmapAnnotation(
  row_data[keep_rows, ],
  which = "row", show_annotation_name = TRUE
)

new_heatmap + new_annots
```

In `functionalheatmap` the whole dataset is passed into a pipeline. You can
filter the contents of either the heatmap body or the row-data and it will
match everything up fine.

The input data looks slightly different than the matrix that is passed to
ComplexHeatmap. We make `tidy` versions of the input data so that tidyverse
verbs (filter / select etc) can be applied to them.

```{r}
tidy_body_df <- body_data %>%
  tibble::as_tibble(rownames = "feature_id") %>%
  tidyr::gather(key = "sample_id", value = "fitted_value", -feature_id)

tidy_row_df <- tibble::rownames_to_column(row_data, "feature_id")
```

```{r}
# we've pulled the matrix indices into a data-frame
head(tidy_body_df)
```

To plot the original heatmap:

```{r}
list(
  body_data = tidy_body_df,
  row_data = tidy_row_df
) %>%
  setup_heatmap() %>%
  format_heatmap(
    cluster_columns = FALSE
  ) %>%
  annotate_heatmap(
    row_annotations = c("foo", "bar"),
    row_dots = list(show_annotation_name = TRUE)
  ) %>%
  plot_heatmap()
```

... and to plot the not-foos, do this:

```{r}
list(
  body_data = tidy_body_df,
  row_data = dplyr::filter(tidy_row_df, !foo)
) %>%
  setup_heatmap() %>%
  format_heatmap(
    cluster_columns = FALSE
  ) %>%
  annotate_heatmap(
    row_annotations = c("foo", "bar"),
    row_dots = list(show_annotation_name = TRUE)
  ) %>%
  plot_heatmap()
```

Or indeed, write a pipeline to return the heatmap data-structure:

```{r}
heatmap_pipeline <- function(x) {
  setup_heatmap(x) %>%
    format_heatmap(
      cluster_columns = FALSE
    ) %>%
    annotate_heatmap(
      row_annotations = c("foo", "bar"),
      row_dots = list(show_annotation_name = TRUE)
    )
}
```

... and then pump your easily filtered dataset into it

```{r}
list(
  body_data = tidy_body_df,
  row_data = dplyr::filter(tidy_row_df, !foo)
) %>%
  heatmap_pipeline() %>%
  plot_heatmap()
```

Using [`polyply`](https://github.com/russHyde/polyply) you can get away with
having just a single heatmap dataset. I won't go into how that works, but the
above code would look like:

```
# -- not ran -- #

# globally-useful heatmap dataset:
library(polyply)
heat_frames <- poly_frame(
  body_data = tidy_body_df,
  row_data = tidy_row_df
)

heat_frames %>%
  # filter based on `foo` in `row_data`
  activate(row_data) %>% filter(!foo) %>%
  # run the functionalheatmap pipeline and plot
  heatmap_pipeline() %>%
  plot_heatmap()
```

In short, `functionalheatmap` allows me to abstract away the functions involved
in making and plotting a heatmap using `ComplexHeatmap`. This has helped
deduplicate and de-`temp-variable`-enate some of my analysis scripts
considerably.

# Use of `mockery::mock()` in `functionalheatmap` tests

An aim of `functionalheatmap` was to be able to incrementally change the
formatting of a heatmap[^1]. For example, if you had already passed some data
through the `heatmap_pipeline` described above, but on calling `plot_heatmap`
you decided the title needed changing, or the body-colours were all wrong
you should be able to do this:

```
# -- not ran -- #
heatmap1 <- blep %>% blep %>% blep %>% heatmap_pipeline()

plot_heatmap(heatmap1)

# YUCK!

heatmap2 <- format_heatmap(heatmap1, na_col = "purple", some_other_formats ...)

plot_heatmap(heatmap2)

# LESS YUCK!
```

... and the formatting you provide should add to or supplant the existing
formatting arguments.

All these formatting flags are stored in the data-structure that is passed
around by the functions in `functionalheatmap`. We still use the Heatmap
plotting code from ComplexHeatmap - we only pass `Heatmap()` the formatting
flags in the final `plot_heatmap` step.

So you can set the formatting flags in one function and they exert their effect
when used in another function.

At one point while writing this package I was unsure how to go about testing
that formatting-flags that were set in `format_heatmap` were doing what they
should. I don't even know _how_ to unit (or otherwise) test the output of a
plotting function. So all I wanted to know was that when a formatting flag is
set in the `format_heatmap` call, that flag is set in the subsequent call to
`Heatmap()` made by `plot_heatmap`.

So I found out about the package
[`mockery`](https://cran.r-project.org/web/packages/mockery/index.html).

The idea is pretty cool. I'd heard of mocks and stubs in passing but had never
used them before. Basically you create a 'mock-object' that behaves _in some
way_ like another object (which we'll name the 'model' cos I was a genetics
student [once](https://en.wikipedia.org/wiki/Mimicry)) and inject that
mock-object into your code. Suppose your doing this within a unit-test, then at
that point where the test-code should use the model-object you actually call
the mock-object.

In `R`, functions are objects, so we can create mock functions. All those
filthy side-effecting functions that touch databases or plot figures or modify
stuff out-of-environment or write to files: they could be replaced with mock
functions in your test code.

In the use-case described above:

- we have a heatmap-data object,

- it is passed through the function `format_heatmap`

    - which sets the `na_col` to `"purple"` for some god-forsaken reason

- then the output from that is passed to `plot_heatmap`

    - and we want to know whether the `na_col` is `"purple"` when
    `plot_heatmap` calls `ComplexHeatmap::Heatmap()`

Evidently, the 'filthy side-effecting' function here is `Heatmap()`. So we will
make a mock `Heatmap` function to help test this use-case.

# Using `mockery` to mock a function

We use `mock` to define a mock function:

```{r}
mock_fn <- mock(2)
```

As set-up, `mock_fn` can be used once (since there's only one argument and we
haven't played with it's `cycle` argument (not discussed)) and when it is
called it will return the value 2.

Then we use `testthat::with_mock` to both

- state which function we want to replace with our mock-function `mock_fn`;

- and write an expression wherein the model-function would be called

Here we want to replace the model-function `Heatmap` and we want to check what
happens when `Heatmap()` is called within `plot_heatmap`. The call to
`with_mock` should look like this:

```
# -- not ran -- #
testthat::with_mock(
  # replace the model-function `Heatmap` with the mock-function `mock_fn`
  Heatmap = mock_fn, {
    # define an expression wherein the model-function would normally be called
    plot_heatmap(some_input_data)
  },
  # you want to do the replacement in the ComplexHeatmap namespace
  .env = "ComplexHeatmap"
)
```

Let's make some input data so we can check everything works as expected:

```{r}
my_heatmap_data <- list(body_data = tidy_body_df) %>%
  setup_heatmap() %>%
  format_heatmap(na_col = "purple")
```

If everything works, the `na_col` parameter in `Heatmap()` should be set to
`purple`.

```{r}
testthat::with_mock(
  Heatmap = mock_fn, {
    plot_heatmap(my_heatmap_data)
  },
  .env = "ComplexHeatmap"
)
```

The value 2 is returned because Heatmap() is the last thing evaluated in
plot_heatmap(), and our mock-function was set-up to return 2.

So what do we test?

The mock-function we set up actually stores some information about the
function call and the argument values used when it was called in place of
Heatmap().

```{r}
mock_args(mock_fn)
```

```{r}
mock_calls(mock_fn)
```

So in an automated test, we can check that `na_col` was set to `purple` as
follows:

```{r}
heatmap_args <- mock_args(mock_fn)
```

```{r, error=TRUE}
testthat::expect_true(
  "na_col" %in% names(heatmap_args[[1]]) &&
    heatmap_args[[1]]$na_col == "orange"
)
```

Ack! That's what happens when your test fails. Sorry, `na_col` was supposed to
be purple:

```{r}
testthat::expect_true(
  "na_col" %in% names(heatmap_args[[1]]) &&
    heatmap_args[[1]]$na_col == "purple"
)
```

Silence is golden.

There's some neater `expect_*` functions for working with mocks within mockery
(`expect_called`, `expect_call`, `expect_args`). They didn't quite hit the mark
for the test I needed though.

So I found `mockery` really helpful. And now, thinking back, I feel pretty bad
about all those tests I once wrote that hit the EBI database servers .....

[^1]: still working on that
